Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks.

### 2020-06-17 main.Rmd
typos

Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress). We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4  to 97.4, for the four fungal models we observed accuracy in the range  60.5 to 84.2 and for the four oomycete models we saw accuracy from 64.7 to 82.3, reported in Fig \@ref(fig:overallmodel). All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.

@@

Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range 0.8 to 0.88, see Figure \@ref(fig:confusionmodelcorrs). CNN-GRU and LSTM-Embedding showed identical prediction sets. We saw similar levels of correlation  in the CNN-LSTM, GRU-Embedding and CNN-GRU fungal sequence trained model, in the range 0.79 to 0.9 ; though there was a significantly lower range of correlations with the LSTM-Embedding which were in the range 0.36 to 0.51. The models trained on oomycete sequences all showed this lower range of correlations, in the range 0.30 to 0.65. The higher correlation across bacterial trained models is likely from a mixture of the larger training set size and a greater uniformity of the sequences themselves. For the fungal sequence trained models we can see that the LSTM-Embedding model does not perform as well as the others. The oomycete sequence trained models all show a lower range correlation reflecting the likely less uniform and smaller training set. It is clear that, particularly for the fungal and oomycete models, each architecture is capturing separate aspects of the sequences and classifying on those with slightly varying levels of success.

@@

We examined the usefulness of combining the predictions of the different model architectures using an ensemble function that takes the vectors of classifications of each model architecture as input. We performed the classification of the hold-out test data set using the ensembled models and the results can be seen in Figure \@ref(fig:ensembleresults). With the models trained in bacterial sequences we saw an increase in classification accuracy over the best model, up to 0.99 for both ensemble functions. However, with the fungal and oomycete models we saw decreases relative to the best single model in both cases due to the higher accuracy of the CNN-LSTM model being diluted by the combined inacuracy of the other model architectures. Examining the overlaps in classifications between the CNN-LSTM and CNN-GRU/LSTM-Embedding respectively showed that the two lesser performing models were not simply predicting subsets of the CNN-LSTM model, in both cases the lesser models were able to identify three effectors correctly that were missed by the generally stronger models. This indicates that the weaker models may be classifying on some patterns missed by the CNN-LSTM model.

@@

Given the accuracies of the above we selected the ensemble bacterial model and the CNN-LSTM fungal and oomycete models to evaluate the performance of our models against widely used effector identification software. We compared against predictions from the bacterial effector prediction programs DeepT3 [@deept3] and EffectiveT3 classification module for plant-associated bacteria 1.0.1 [@effectivet3], the fungal effector prediction programs EffectorP 1.0 and 2.0 [@effectorp, @effectorp2] and the oomycete effector predictor EffectR [@effectR]. Each comparison was carried out using the respective hold-out test sequence set for each taxon. For all taxon we saw greater Accuracy and F1 scores from our models than the established software, as shown in Figure \@ref(fig:vsothersmetrics). This was particularly marked in the F1 score, which incorporates a measure of the incorrect calls. Absolute improvements were up to 15 %  in bacterial sequences, 15 % in fungal sequences and 20 %  in the oomycete sequences. The confusion matrices in Figure \@ref(fig:vsotherscfmat) show that accuracy and F1 score was compromised in all the established tools by the tendency of them all to misclassify true effectors as not effectors.  All the established software classifiers we tested show lower sensitivity than the models we have developed here.

@@

  mkplot <- function(df, which_tax ) {
    df$Tool <- factor(df$Tool, levels = c("cnn_lstm", "ens_weight", "Effective T3 All", "Effective T3 Plants", "Deep T3", "EffectorP", "EffectorP 2", "effectR"))
    p <- df %>% filter(Taxon == which_tax) %>% 
     ggplot() +
    aes(Tool, `Proportion Found`) +
    geom_bar(stat= 'identity', fill = rgb(52/255, 106/255, 141/255), colour = rgb(52/255, 106/255, 141/255), alpha=0.5) +
    facet_grid(  ~ `Data Set`, scales = "free") + 
      theme_bw() +
      scale_color_viridis_d()
  }

>>>>

Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress). We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4  to 97.4, for the four fungal models we observed accuracy in the range  60.5 to 84.2 and for the four oomycete models we observed accuracy from 64.7 to 82.3, reported in Fig \@ref(fig:overallmodel). All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.

@@

Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range 0.8 to 0.88, see Figure \@ref(fig:confusionmodelcorrs). CNN-GRU and LSTM-Embedding showed identical prediction sets. We observed similar levels of correlation  in the CNN-LSTM, GRU-Embedding and CNN-GRU fungal sequence trained model, in the range 0.79 to 0.9 ; though there was a significantly lower range of correlations with the LSTM-Embedding which were in the range 0.36 to 0.51. The models trained on oomycete sequences all showed this lower range of correlations, in the range 0.30 to 0.65. The higher correlation across bacterial trained models is likely from a mixture of the larger training set size and a greater uniformity of the sequences themselves. For the fungal sequence trained models we can see that the LSTM-Embedding model does not perform as well as the others. The oomycete sequence trained models all show a lower range correlation reflecting the likely less uniform and smaller training set. It is clear that, particularly for the fungal and oomycete models, each architecture is capturing separate aspects of the sequences and classifying on those with slightly varying levels of success.

@@

We examined the usefulness of combining the predictions of the different model architectures using an ensemble function that takes the vectors of classifications of each model architecture as input. We performed the classification of the hold-out test data set using the ensembled models and the results can be seen in Figure \@ref(fig:ensembleresults). With the models trained in bacterial sequences we observed an increase in classification accuracy over the best model, up to 0.99 for both ensemble functions. However, with the fungal and oomycete models we observed decreases relative to the best single model in both cases due to the higher accuracy of the CNN-LSTM model being diluted by the combined inacuracy of the other model architectures. Examining the overlaps in classifications between the CNN-LSTM and CNN-GRU/LSTM-Embedding respectively showed that the two lesser performing models were not simply predicting subsets of the CNN-LSTM model, in both cases the lesser models were able to identify three effectors correctly that were missed by the generally stronger models. This indicates that the weaker models may be classifying on some patterns missed by the CNN-LSTM model.

@@

Given the accuracies of the above we selected the ensemble bacterial model and the CNN-LSTM fungal and oomycete models to evaluate the performance of our models against widely used effector identification software. We compared against predictions from the bacterial effector prediction programs DeepT3 [@deept3] and EffectiveT3 classification module for plant-associated bacteria 1.0.1 [@effectivet3], the fungal effector prediction programs EffectorP 1.0 and 2.0 [@effectorp, @effectorp2] and the oomycete effector predictor EffectR [@effectR]. Each comparison was carried out using the respective hold-out test sequence set for each taxon. For all taxa we observed greater Accuracy and F1 scores from our models than the established software, as shown in Figure \@ref(fig:vsothersmetrics). This was particularly marked in the F1 score, which incorporates a measure of the incorrect calls. Absolute improvements were up to 15 %  in bacterial sequences, 15 % in fungal sequences and 20 %  in the oomycete sequences. The confusion matrices in Figure \@ref(fig:vsotherscfmat) show that accuracy and F1 score was compromised in all the established tools by the tendency of them all to misclassify true effectors as not effectors.  All the established software classifiers we tested show lower sensitivity than the models we have developed here.

@@

  mkplot <- function(df, which_tax ) {
    df$Tool <- factor(df$Tool, levels = c("cnn_lstm", "ens_weight", "Effective T3 All", "Effective T3 Plants", "Deep T3", "EffectorP", "EffectorP 2", "effectR"))
    p <- df %>% filter(Taxon == which_tax) %>% 
     ggplot() +
    aes(Tool, `Proportion Found`) +
    geom_bar(stat= 'identity', fill = rgb(52/255, 106/255, 141/255), colour = rgb(52/255, 106/255, 141/255)) +
    facet_grid(  ~ `Data Set`, scales = "free") + 
      theme_bw() +
      scale_color_viridis_d()
  }
